From Reversible Programs to \\ Univalent Universes and Back

### 2021-11-11 Pi/paper/discussion.tex
Fix headlines and captions
Citations

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible computing based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels}. In this work, we have established this
correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we discuss some broader
related work.

\begin{figure}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            \rowcolor{lightgray} $\PiLang$ & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline
            Types                          & Natural numbers               & Finite sets & 0-cells \tabularnewline
            \hline
            1-combinators                  & Generators of $\Sn$           & Bijections  & 1-paths \tabularnewline
            \hline
            2-combinators                  & Relations of $\Sn$            & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \caption{Summary of the established correspondences}
    \label{fig:table-summary}
\end{figure}

@@

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
    {homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
describes an explicit algorithm for producing normal forms. It could provide an alternative to our rewriting system. Other
encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
string diagrams, proved either difficult to formalise in type theory or difficult to relate directly to the primitive type
isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
making proving correctness and termination intractable.

@@

%%%%%%%%%%%%%%
\paragraph{Reversible Programming Languages} The practice of programming languages is replete with \emph{ad hoc}
instances of reversible computations: database transactions, mechanisms for data provenance, checkpoints, stack and
exception traces, logs, backups, rollback recoveries, version control systems, reverse engineering, software
transactional memories, continuations, backtracking search, and multiple-level undo features in commercial applications.
In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of
reversibility. But intensive research in full-fledged reversible models of computations and reversible programming
languages was only sparked by the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,Hey:1999:FCE:304763,fredkin1982conservative},
and by the potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}. The early developments of
reversible programming languages started with a conventional programming language, e.g., an extended $\lambda$-calculus,
and either (i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or (ii) imposed constraints on the control
flow constructs to make them reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.  More modern approaches recognize that
reversible programming languages require a fresh approach and should be designed from first principles without the
detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}. The version of
$\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It would be
interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
types~\cite{jamesInformationEffects2012,rc2011}, or negative and fractional
types~\cite*{chenFractionalTypes2020,chenComputationalInterpretationCompact2021}.

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible computing based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{caretteEmbracingLawsPhysics2021}. In this work, we
have established this correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we
discuss some broader related work.

\begin{figure}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            \rowcolor{lightgray} $\PiLang$ & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline
            Types                          & Natural numbers               & Finite sets & 0-cells \tabularnewline
            \hline
            1-combinators                  & Generators of $\Sn$           & Bijections  & 1-paths \tabularnewline
            \hline
            2-combinators                  & Relations of $\Sn$            & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \caption{Summary of the established correspondences TODO: Move above}
    \label{fig:table-summary}
\end{figure}

@@

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
\citet{hiverCoqCombiCoxeterPresentation2021} describes an explicit algorithm for producing normal forms. It could
provide an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult to formalise in type theory
or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
intractable.

@@

%%%%%%%%%%%%%%
\paragraph{Reversible Programming Languages} The practice of programming languages is replete with \emph{ad hoc}
instances of reversible computations: database transactions, mechanisms for data provenance, checkpoints, stack and
exception traces, logs, backups, rollback recoveries, version control systems, reverse engineering, software
transactional memories, continuations, backtracking search, and multiple-level undo features in commercial applications.
In the early nineties, \citet{bakerNREVERSALFortuneThermodynamics1992,bakerLivelyLinearLisp1992} argued for a
systematic, first-class, treatment of reversibility. But intensive research in full-fledged reversible models of
computations and reversible programming languages was only sparked by the discovery of deep connections between physics
and computation~\cite{landauerIrreversibilityHeatGeneration1961, peresReversibleLogicQuantum1985,
    toffoliReversibleComputing1980, bennettFundamentalPhysicalLimits1985, frankReversibilityEfficientComputing1999,
    heyFeynmanComputationExploring1999, fredkinConservativeLogic1982}, and by the potential for efficient quantum
computation~\cite{feynmanSimulatingPhysicsComputers1982}. The early developments of reversible programming languages
started with a conventional programming language, e.g., an extended $\lambda$-calculus, and either (i) extended the
language with a history mechanism~\cite{vantonderLambdaCalculusQuantum2004,klugeReversibleSECD2000,
    huelsbergenLogicallyReversibleEvaluator1996,danosReversibleCommunicatingSystems2004}, or (ii) imposed constraints on the
control flow constructs to make them reversible~\cite{yokoyamaReversibleProgrammingLanguage2007}. More modern approaches
recognize that reversible programming languages require a fresh approach and should be designed from first principles
without the detour via conventional irreversible languages~\cite{yokoyamaPrinciplesReversibleProgramming2008,
    muInjectiveLanguageReversible2004, abramskyStructuralApproachReversible2005, dipierroReversibleCombinatoryLogic2006}.
The version of $\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It
would be interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with
recursive types~\cite{jamesInformationEffects2012,bowmanDaggerTracedSymmetric2011}, or negative and fractional
types~\cite*{chenFractionalTypes2020,chenComputationalInterpretationCompact2021}.

### 2021-11-11 Pi/paper/finite.tex
Fix headlines and captions
Citations

\subsection{Presenting the symmetric group}

@@

\subsection{Free groups}

@@

\subsection{Group presentations}

@@

Lehmer codes are known in Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called
"subexcedant sequences", or "factoriadics". They can be written as a decimal number in the factorial number system, or
as a tuple encoding the digits~\cite*{knuthArtComputerProgramming1997,laisantNumerationFactorielleApplication1888}. This
gives a convenient way of representing permutations on a computer, partly because they are
bitwise-optimal~\cite{bergerTeachingOrdinalPatterns2019a}.

@@

\subsection{Running Lehmer codes}

Finally, it is time to complete our goal of characterising the symmetric group. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$~\footnote{Note that the indices for the type of permutations
  are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
this proof is borrowed from~\citet{Molzer-cubical}.

>>>>

\subsection{Presenting the Symmetric Group}

@@

\subsection{Free Groups}

@@

\subsection{Group Presentations}

@@

Lehmer codes are known in Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called
"subexcedant sequences", or "factoriadics". They can be written as a decimal number in the factorial number system, or
as a tuple encoding the digits~\cite*{knuthArtComputerProgramming1997,laisantNumerationFactorielleApplication1888}. This
gives a convenient way of representing permutations on a computer, partly because they are
bitwise-optimal~\cite{bergerTeachingOrdinalPatterns2019}.

@@

\subsection{Running Lehmer Codes}

Finally, it is time to complete our goal of characterising the symmetric group. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$~\footnote{Note that the indices for the type of permutations
are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
this proof is borrowed from~\citet{molzerCubicalAgdaSimple2021}.

### 2021-11-11 Pi/paper/introduction.tex
Say symmetric group instead of permutation group
More edits
Citations

Consider two programs that implement the same permutation using different sequences of type isomorphisms:
\begin{align*}
      A + (B + C) \isoone
      (A + B) + C & \isoone
      C + (A + B) \isoone
      C + (B + A)
      \\
      A + (B + C) \isoone
      A + (C + B) \isoone
      (A + C) + B & \isoone
      (C + A) + B \isoone
      C + (A + B) \isoone
      C + (B + A)
\end{align*}

@@

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronson_et_al:LIPIcs:2017:8173,1201583}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic datatypes as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012,theseus} whose syntax is inspired by the
sound and complete axiomatisation of type isomorphisms~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the
bicartesian structure, that is, coproducts and products. Since $\PiLang$ programs correspond to type isomorphisms of
finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid (on
zero generators).

It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on zero
generators~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
equational theory for $\PiLang$ that exactly includes all the necessary equations to decide equivalence of $\PiLang$
programs. As conjectured by \citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence
      conditions of symmetric rig groupoids}, answering the conjecture in the positive.

\paragraph{Equivalence by Example} Without going into proofs, let us try to answer the question for the original
pair of examples.

@@

      \[
            \Tree [ {\tiny A} [ {\tiny B} {\tiny C} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny A} [ {\tiny C} {\tiny B} ] ] ~\xrightarrow{\assoclp}~
            \Tree [ [ {\tiny A} {\tiny C} ] {\tiny B} ] ~\xrightarrow[\swapp\phantom{x}]{\phantom{x}\idc}~
            \Tree [ [ {\tiny C} {\tiny A} ] {\tiny B} ] ~\xrightarrow{\assocrp}~
            \Tree [ {\tiny C} [ {\tiny A} {\tiny B} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny C} [ {\tiny B} {\tiny A} ] ] ~
      \]
      \label{fig:example-first-stage}
      \caption{Type isomorphisms as tree transformations}
\end{figure}

@@

      \label{fig:example-braid}
      \caption{Type isomorphisms as braid diagrams}
\end{figure}

The first step is to realise that the use of associativity is uninteresting, and the only steps with interesting
computational content are the swaps. The swaps can be either big or small -- they can happen between leaf nodes, or
between bigger subtrees, respectively.

First, we normalise the types to a list-like representation $A + (B + (C + \zerot))$, where $\zerot$ is the empty type.
In this representation, each type is identified with its index, $A$ has index 0, $B$ has index 1, and $C$ has index 2.
Then, we need to compile each primitive isomorphism to a list of adjacent transpositions, and then compose by appending
the lists. A small swap is trivially implemented as a single adjacent transposition. To compile a big swap into adjacent
transpositions, we need to traverse the subtrees in-order and recursively swap elements from the left by transposing
them across the ones in the middle -- this generates a large number of adjacent transpositions. For the two programs, we
get the following two lists: $[1,0,1,1,1]$ (\cref{fig:cone-braid}) and $[1,0,1]$ (\cref{fig:ctwo-braid}), where the
number $k$ encodes a transposition of elements at indices $k$ and $k+1$.

Swapping is symmetric, that is, swapping two elements and immediately swapping them back produces the same permutation.
The first list contains such redundant operations, so to eliminate these, we have to normalise the lists, which we will
do by setting up reduction relations and an appropriate rewriting system. This system will normalise both lists to
$[0,1,0]$ (\cref{fig:norm-braid}), which is lexicographically the smallest list that corresponds to this permutation.
Since both programs have the same normal forms, they're equivalent!

The crux of designing an equational theory for $\PiLang$ relies on the choice of relations we use to design this
rewriting system. First, there should be enough equations relating the programs to their normalised forms, and second,
there should be enough equations corresponding to the reduction rules. We will show that these correspond to the
coherence conditions for symmetric rig groupoids.

From the normalised list $[0,1,0]$, we can construct a permutation on a list of 3 elements $[a,b,c]$ as follows. We think
of it as insertion-sorting the elements of the list. Starting from the list $[a, b, c]$, we first insert $b$ at the
right place -- by applying transposition $0$, we get the list $[b, a, c]$. Then, element $c$ is inserted in the right
place by applying transpositions $1$ and $0$ -- as a result, we get the desired permutation $[c, b, a]$. Notice that we
could specify a more compact way of describing this process, since the key information was only how many shifts we
needed to apply to an element. We could describe this procedure using a code $(0, 1, 2)$, which says how many inversions
to apply to elements $a$, $b$, and $c$, respectively.

@@

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
state our result as a Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and
Symmetric Rig Groupoids, using which we can build a toolbox of technical devices for reasoning about reversible
circuits.

@@

      \item We start in~\cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to
            serve as running examples throughout the paper.
      \item In~\cref{sec:pi}, we introduce the two-level language
            $\PiLang$~\cite{jamesInformationEffects2012,caretteComputingSemiringsWeak2016} and illustrate how to write
            reversible circuits and their equivalences using 1-combinators and 2-combinators respectively. We give a
            semantic account of the language by translating each level-1 program to a bijection between finite sets, and
            verifying that programs identified by level-2 constructs denote the same bijection.
      \item \cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
            Homotopy Type Theory and Univalent Foundations. We define and characterise the notion of a univalent subuniverse, and construct
            $\UFin$ as a univalent subuniverse which classifies all finite types. We establish that paths in $\UFin$ are
            families of loops on finite sets of specified cardinality, given by $\Aut[\Fin[n]]$, which produces the
            permutation group on $\Fin[n]$.
      \item In~\cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group

@@

            stages. First we define a subset $\PiPlusLang$ of the language which only includes the additive monoidal

>>>>

Consider two programs that implement the \emph{same} permutation using different sequences of type isomorphisms:
\begin{align*}
      A + (B + C) \leftrightarrow
      (A + B) + C & \leftrightarrow
      C + (A + B) \leftrightarrow
      C + (B + A)
      \\
      A + (B + C) \leftrightarrow
      A + (C + B) \leftrightarrow
      (A + C) + B & \leftrightarrow
      (C + A) + B \leftrightarrow
      C + (A + B) \leftrightarrow
      C + (B + A)
\end{align*}

@@

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronsonClassificationReversibleBit2017}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic data types, as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012, jamesTheseusHighLevelLanguage2014}. The
syntax of $\PiLang$ is inspired by the sound and complete axiomatisation of type isomorphisms of lambda
calculi~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the bicartesian structure, that is, product and sum
types. $\PiLang$ programs correspond to type isomorphisms of finite types, and our observation is that the syntax of
$\PiLang$ is a presentation of the free symmetric rig groupoid (on zero generators).

It is folklore that the groupoid of finite sets and bijections is the free symmetric rig groupoid on zero
generators~\cite{laplazaCoherenceDistributivity1972, kellyCoherenceTheoremsLax1974, baezFiniteSetsFeynman2001}. Our main
result formally establishes this correspondence by giving an equational theory for $\PiLang$ that includes exactly all
the necessary equations to decide equivalence of $\PiLang$ programs. As conjectured
by~\citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence conditions of symmetric
      rig groupoids}, answering the conjecture in the positive.

\paragraph{Equivalence by Example} Without going into details of proofs, let us try to answer the question for the
original pair of examples.

@@

      \[
            \Tree [ {\tiny A} [ {\tiny B} {\tiny C} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny A} [ {\tiny C} {\tiny B} ] ] ~\xrightarrow{\assoclp}~
            \Tree [ [ {\tiny A} {\tiny C} ] {\tiny B} ] ~\xrightarrow[\swapp\phantom{x}]{\phantom{x}\idc}~
            \Tree [ [ {\tiny C} {\tiny A} ] {\tiny B} ] ~\xrightarrow{\assocrp}~
            \Tree [ {\tiny C} [ {\tiny A} {\tiny B} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny C} [ {\tiny B} {\tiny A} ] ] ~
      \]
      \caption{Type isomorphisms as tree transformations}
      \label{fig:example-first-stage}
\end{figure}

@@

      \label{fig:example-braid}
      \caption{Type isomorphisms as braiding diagrams}
\end{figure}

The first step is to realise that the use of associativity is uninteresting, and the only steps with interesting
computational content are the swaps. The swaps can be either \emph{small} or \emph{big} -- they can happen between leaf
nodes, or between bigger subtrees, respectively, as shown in~\cref{fig:example-first-stage}.

First, we normalise the types to a list-like representation $A + (B + (C + \zerot))$, where $\zerot$ is the empty type.
In this representation, each type is identified with its index, $A$ has index 0, $B$ has index 1, and $C$ has index 2.
Then, we compile each primitive isomorphism to a list of adjacent transpositions, which are composed by appending the
lists. A small swap is trivially implemented as a single adjacent transposition. To compile a big swap into adjacent
transpositions, we traverse the subtrees in-order and recursively swap elements from the left by transposing them across
the ones in the middle -- this generates a large number of adjacent transpositions. For the two programs, we get the
following two lists: $[1,0,1,1,1]$ (\cref{fig:cone-braid}) and $[1,0,1]$ (\cref{fig:ctwo-braid}), where the number $k$
encodes a transposition of elements at indices $k$ and $k+1$.

Swapping is symmetric, that is, swapping two elements and immediately swapping them back produces the same permutation.
The first list contains such redundant operations, so to eliminate these, we will normalise the lists, by setting up
reduction relations and an appropriate rewriting system. This system normalises both lists to $[0,1,0]$
(\cref{fig:norm-braid}), which is lexicographically the smallest list that corresponds to this permutation. Since both
programs have the same normal form, they're equivalent!

The crux of designing an equational theory for $\PiLang$ relies on the choice of relations we use to design this
rewriting system. First, there should be enough equations to compile programs to their normalised forms, that is,
sequences of adjacent transpositions, and second, there should be enough equations to reduce them to a normal form, that
is, the reduced list of transpositions. We will show that these equations correspond to the coherence conditions for
symmetric rig groupoids.

From the normalised list $[0,1,0]$, we can recover a permutation on a list of 3 elements $[a,b,c]$ as follows. We think
of it as insertion-sorting the elements of the list. Starting from the list $[a, b, c]$, we first insert $b$ at the
right place -- by applying transposition $0$, we get the list $[b, a, c]$. Then, element $c$ is inserted in the right
place by applying transpositions $1$ and $0$ -- as a result, we get the desired permutation $[c, b, a]$. Notice that we
could specify a more compact way of describing this process, since the key information was only how many shifts we
needed to apply to an element. We could describe this procedure using a code $(0, 1, 2)$, which says how many inversions
to apply to elements $a$, $b$, and $c$, respectively.

@@

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
state our result as a Curry-Howard-Lambek correspondence for Reversible
Logic~\cite{sparksSuperstructuralReversibleLogic2014}, Reversible Programming
Languages~\cite{jamesInformationEffects2012}, and Symmetric Rig Groupoids~\cite{laplazaCoherenceDistributivity1972,
kellyCoherenceTheoremsLax1974, caretteComputingSemiringsWeak2016}, using which we can build a toolbox of technical
devices for reasoning about reversible circuits.

@@

      \item We start in~\cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit
            framework~\cite{aleksandrowiczQiskitOpensourceFramework2019} to serve as running examples throughout the
            paper.
      \item In~\cref{sec:pi}, we introduce the two-level language $\PiLang$~\cite{caretteComputingSemiringsWeak2016} and
            illustrate how to write reversible circuits and their equivalences using 1-combinators and 2-combinators
            respectively. We give a semantic account of the language by translating each level-1 program to a bijection
            between finite sets, and verifying that programs identified by level-2 constructs denote the same bijection.
      \item \cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and bijections, in
            Homotopy Type Theory and Univalent Foundations (HoTT/UF). We define and characterise the notion of a
            univalent subuniverse, and construct $\UFin$ as a univalent subuniverse which classifies all finite types.
            We establish that paths in $\UFin$ are families of loops on finite sets of specified cardinality, given by
            $\Aut[\Fin[n]]$, which produces the group of permutations on $\Fin[n]$.
      \item In~\cref{sec:finite}, we proceed to give a presentation of this group, as the symmetric group

@@

            stages. First, we define a subset $\PiPlusLang$ of the language which  includes the additive monoidal

### 2021-11-11 Pi/paper/reversible1.tex
Citations

The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the
hardware platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an
``assembly-level'' programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical
model based on permutations of finite sets provides a richer algebraic structure, using which we describe a reversible
programming language in this section.

@@

% A specialised language for programming with permutations on finite sets can be built using ideas going back to Kelly,
% Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
% rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
% Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
% language and the algebraic closure operators become composition operators in the programming language. Once equipped
% with a notion of values and types, we get the following syntactic presentation of a programming language for
% permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig
groupoids was established by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:
\medskip
\leveltwoblockone{}

@@

For combinators, we show explicitly how values reduce along each combinator, similar to a big-step operational
semantics~\cite{chenComputationalInterpretationCompact2021,theseus}.

>>>>

The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the
hardware platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an
``assembly-level'' programming abstraction (e.g., \verb|qasm|). As motivated by
\citet{lafontAlgebraicTheoryBoolean2003}, a mathematical model based on permutations of finite sets provides a richer
algebraic structure, using which we describe a reversible programming language in this section.

@@

% A specialised language for programming with permutations on finite sets can be built using ideas going back to Kelly,
% Laplaza, and Mac Lane~\cite{laplazaCoherenceDistributivity1972,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
% rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
% Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
% language and the algebraic closure operators become composition operators in the programming language. Once equipped
% with a notion of values and types, we get the following syntactic presentation of a programming language for
% permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig groupoids was established
by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this semantics, \emph{coherence
conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplazaCoherenceDistributivity1972,caretteComputingSemiringsWeak2016}, were collected in a second
level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for
appropriate $c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same
bijection. For example, we have the following level-2 combinators dealing with associativity:
\medskip
\leveltwoblockone{}

@@

For combinators, we show explicitly how values reduce along each combinator, similar to a big-step operational
semantics~\cite{chenComputationalInterpretationCompact2021,jamesTheseusHighLevelLanguage2014}.

### 2021-11-11 Pi/paper/ufin.tex
Fix headlines and captions
Citations

\subsection{Rig structure}
\label{subsec:rig}

@@

\begin{theorem}
  $\UFin$ has two symmetric monoidal structures, the additive and multiplicative ones, given by $(F_0, \sqcup)$ and
  $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the braiding
  isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
  for symmetric monoidal categories~\cite{maclaneNaturalAssociativityCommutativity1963}, that is, the triangle,
  pentagon, and hexagon identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. The multiplicative
  structure distributes over the additive structure and satisfies the Laplaza coherence conditions for rig
  categories~\cite{laplaza72}.
\end{theorem}

>>>>

\subsection{Rig Structure}
\label{subsec:rig}

@@

\begin{theorem}
  $\UFin$ has two symmetric monoidal structures, the additive and multiplicative ones, given by $(F_0, \sqcup)$ and
  $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the braiding
  isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
  for symmetric monoidal categories~\cite{maclaneNaturalAssociativityCommutativity1963}, that is, the triangle,
  pentagon, and hexagon identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. The multiplicative
  structure distributes over the additive structure and satisfies the Laplaza coherence conditions for rig
  categories~\cite{laplazaCoherenceDistributivity1972}.
\end{theorem}
